---
title: "Skills in the BC labour market"
author: "Ermias Afeworki, Richard Martin"
output:
  bookdown::word_document2:
    fig_caption: true
    fig_caption_location: bottom
    toc: false           # optional: add if you want a TOC
    number_sections: false # optional: number sections or not
    reference_docx: null  # optional: specify your Word template here
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  error = FALSE
)

if (file.info(here::here("out","richs_objects.rds"))$mtime < file.info("01_source_rich.R")$mtime) {
  stop("richs_objects.rds is stale — rerun 01_source_rich.R")
}
library(tidyverse)
ro <- readr::read_rds(here::here("out", "richs_objects.rds"))

all_occs_chart <- read_rds(here::here("out", "all_occs_chart.rds"))
plot_upper <- read_rds( here::here("out", "plot_upper.rds"))
plot_lower <- read_rds(here::here("out", "plot_lower.rds"))
plot_dist <- read_rds(here::here("out", "plot_dist.rds"))

library(readxl)
library(here)

# Read a specific worksheet
skill_clusters <- read_excel(here("out", "skill_clusters_data.xlsx"), sheet = "skill_clusters_tbl") |> 
  mutate(across(everything(), ~ ifelse(is.na(.), "", .))) # Replace NA with blank space


```

`r format(lubridate::now(), "%Y-%m-%d %H:%M:%S")#timestamp handy while developing`

In order for the BC economy to thrive it is essential that the skills of the workforce match the needs of employers. However, the labour market is constantly evolving both in terms of occupational composition as well as the skills required for each specific occupation. In this note we focus exclusively on composition changes, and make the assumption that occupational skill profiles are constant. This assumption is supported by data frequency: employment by occupation is available (and variable) monthly in the Labour Force Survey, whereas the O\*NET skills data uses a rolling update schedule, completing a full refresh of all occupations every five years.

The goal is to identify the skills that are becoming more important in the labour market. We begin by quantifying the broad shifts in the labour market by looking at how the TEER composition has changed historically, and is forecast to change over the 10 years covered by the Labour Market Outlook. Once we have established the broad shifts in the labour market, we then compare the current skill profiles of High Opportunity Occupations to the rest of the labour market. Finally we look at how the distribution of skills is changing over time, both historically (2015-2024) and in the future (2025-2035).

## A first look: breakdown by TEER

We can get a preliminary idea of the changes in the labour market by looking at the breakdown of employment by TEER (Training, Education, Experience and Responsibilities) categories. The TEER system is a classification system used in Canada to categorize occupations based on the training, education, experience, and responsibilities required for each job. If the TEER composition of the labour market is changing over time it is a clear indication that the required skills likewise changing.

```{r, fig1, fig.cap="Employment by TEER groupings", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
ro$emp_teer_plt 
```

## High opportunity skills

In recognition of the dynamic nature of the labour market, the Labour Market Outlook identifies certain occupations as being High Opportunity Occupations on the basis of seven economic indicators. These indicators can be grouped into four categories, which are based on current and future labour market conditions. Each occupation is given a score based on its ranking in each of the seven economic indicators, and scores are weighted based on their labour market impact. The occupations are then sorted by score, and the 125 occupations with the highest scores become High Opportunity Occupations.

In what follows we create two weighted average skill profiles, one for the High Opportunity Occupations and one for the rest of the labour market (Figure \@ref(fig:fig2)). It is evident that High Opportunity Occupations require higher skills, with a few exceptions e.g. repairing and equipment maintenance. Using these weighted average skill scores, we investigate the relative (Figure \@ref(fig:fig3)) and absolute (Figure \@ref(fig:fig4)) differences to identify high opportunity skills: the skills that have both relatively and absolutely higher scores in high opportunity occupations: e.g. Science and Operations Analysis. 

```{r fig2, fig.cap='Weighted average skills by HOO vs other', fig.width=9, fig.height=11, dpi=400, out.width="110%"}
ro$skills_plt
```

```{r, fig3, fig.cap="Relative differences", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
ro$diffs_plt1
```

```{r, fig4, fig.cap="Absolute differences", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
ro$diffs_plt2
```

## Mean values and growth rates of skills

An alternative way to investigate the changes in required skills over time is to look at the labour market as a whole and make comparisons across time. We calculate the weighted average skills at various points in time, and then we calculate the compound annual growth rate.  This gives us a sense of how important each skill is currently, and how it is changing over time. The following plot shows the weighted average skill scores in 2024 (Figure \@ref(fig:fig5)) and the compound annual growth rate (CAGR) of these weighted average skill scores from 2015 to 2024 (Figure \@ref(fig:fig6)), and from 2025 to 2035 (Figure \@ref(fig:fig7)).

```{r, fig5, fig.cap="Weighted average skills 2024", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
ro$crossed_plt1
```

```{r, fig6, fig.cap="Historic change", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
ro$crossed_plt2
```

```{r fig7, fig.cap="Future change", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
ro$crossed_plt3 
```

## Skill Clusters

British Columbians are experiencing a shift in the nature of work, driven by factors such as rapid technological advancements, including artificial intelligence, and demographic changes like an aging population. Understanding which occupational skills will be important in the jobs of the future is essential for helping job seekers and employers adapt to the evolving economy and labour market. This information is also valuable for educators and policymakers in designing programs that equip the future workforce with the skills and competencies needed for success over the next ten years.
In previous labour market reports, we analyzed the demand for individual skills (2022 LMO) and occupational clusters based on skill profiles (2023 LMO). In the updated skills and competencies data from O*NET used in this project, we observed that some skills are highly correlated. For example, our analysis shows that the correlation coefficient among the ‘Social’ skills exceeds 0.9. As a result, this report shifts focus from individual skills to clustering skills based on their usage patterns across occupations.

While many clustering techniques are intended to group observations (such as the 11 occupational clusters in the 2023 LMO), we explored several methods for grouping correlated variables, in this case, skills. We adopted the R clustering package ClustOfVar, developed by Chavent et al. (2012), and validated our results using another R package, ClustVarLV, developed by Vigneau and Qannari (2003). While O*NET provides both importance and level ratings for skills and competencies, we focus on importance ratings in this report, as the two measures are highly correlated.

The resulting skill clusters revealed distinct groupings, which we termed analytical, managerial, social, STEM, and technical skill clusters. These clusters offer a more interpretable and actionable framework for understanding future skills demand.

To assess the importance of each skill cluster within occupations, we calculated the average importance score for each cluster. We then applied standardized thresholds to interpret these scores, using O*NET’s anchor-based scale:
•	Important: ≥ 50
•	Moderately Important: 25–49
•	Less Important: < 25

To validate this approach, we compared the distribution of average skill cluster scores with the original O*NET importance ratings for individual skills, as mapped to Canadian occupations. Statistically, the 25th, 50th, and 75th percentiles of the cluster scores were approximately 25.0, 38.6, and 50.7, respectively, closely aligning with the fixed thresholds. Visually, both distributions showed similar peaks near 25 and 50, despite the cluster-level scores being smoother and more compressed due to averaging. This reinforces the validity of our categorization while acknowledging that extreme values are less likely after aggregation (_Figure 8_).



```{r, fig8, fig.cap="Comparison of skill importance distributions", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
plot_dist
```

Next, we linked the average cluster scores to projected job openings, enabling us to identify which skill clusters are most likely to be in demand over the next decade. Finally, we presented results for all occupations and provided disaggregated insights by grouping occupations according to the Training, Education, Experience, and Responsibilities (TEER) framework.

The analytical skills cluster includes foundational abilities such as active listening, reading comprehension, and writing, which tend to be highly rated across many occupations. It also encompasses cognitive skills like critical thinking, complex problem solving, and judgment and decision making, which are similarly valued across diverse roles. Our findings indicate that analytical skills will be essential for success in 66% of job openings over the next ten years and considered somewhat important in the remaining 34%. This suggests that some level of competency in these skills will be required for all job openings during this period (*Figure 9*).

Some level of competency in social and management skills is also required across the board, with these skills rated as either important (56% and 27%, respectively) or somewhat important (44% and 73%, respectively) in all job openings projected over the next ten years. In contrast, STEM and technical skills are rated as important in less than 5% of future job openings (*Figure 9*). This suggests that, from a broad labour market perspective, these skills are considered less critical than other skill clusters. However, they tend to be more specialized and occupation-specific, with their importance becoming more evident when the data is disaggregated by factors such as Training, Education, Experience, and Responsibility (TEER) requirements.



```{r, fig9, fig.cap="All Occupations: Percent of Job Openings 2025", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
all_occs_chart
```

Management occupations (TEER 0) typically demand higher levels of expertise and experience, which is reflected in our analysis: over 95% of job openings in this category require a broad range of essential skills, including analytical, social, and management skills. In occupations requiring a university degree (TEER 1), analytical skills are rated as important in 99% of future job openings. While social and management skills are also essential for this group, the level of proficiency required is somewhat lower compared to that expected in management-specific roles (*Figure 10*).

STEM skills show relatively higher importance in occupations requiring a university degree (TEER 1), and to some extent in TEER 0 occupations, albeit at a lower proficiency level. STEM skills are rated as important in 16% of future job openings that require a university degree, reflecting their relevance in more specialized and knowledge-based occupations(*Figure 10*). In contrast, technical skills are more commonly emphasized in occupations classified under TEER 2 to TEER 5, which typically require college/apprenticeship training or a high school diploma or less. While technical skills are rarely rated as “Important,” they are still considered moderately important in over one-third of job openings within these categories, highlighting their role in supporting practical, hands-on work across a wide range of occupations (*Figure 11*).


```{r, fig10, fig.cap="Management and University", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
plot_upper
```

```{r fig11, fig.cap="College/Apprenticeship and Highschool or less", fig.width=9, fig.height=5.5, dpi=400, out.width="110%"}
plot_lower
```


```{r, echo=FALSE, results='asis'}
cat(rep("\n\n&nbsp;\n\n", 15), sep = "") #hack to force the table onto next page... can't do pagebreaks with bookdown
```


```{r, echo=FALSE, results='asis'}
# Render the table
library(knitr)
kable(skill_clusters, caption = "Skill Cluster Members", booktabs = TRUE)
```

